---
title: "Odds ratios"
lightbox: true
---
```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
import shutup;shutup.please()
exec(open('setup_files/setup.py').read())
```

::: {.callout-tip}
## Learning outcomes

-   Understand how and when to use odds ratios to help interpret logistic regression
-   Know the difference between odds ratios and risk ratios
-   Be able to calculate and discuss/interpret odds ratios
:::

## Libraries and functions

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

### Libraries

```{r}
#| eval: false
library(performance)
library(tidyverse)
library(broom)
```

## Python

### Libraries

```{python}
#| eval: false
# A maths library
import math
import pandas as pd
from plotnine import *
import statsmodels.api as sm
import statsmodels.formula.api as smf
from scipy.stats import *
import numpy as np
```
:::
:::

Introduce & describe dataset

## What is an odds ratio?

## Calculating odds ratios

## Interpretation

The main effect of surgery_type refers to the difference in log-odds relative to the reference level.

The surgery_duration coefficient represents the change in odds per additional minute of surgery (unless you rescale it).

Interaction terms are multiplicative on the odds scale (e.g., modifying the slope of duration depending on surgery type).

### Categorical predictor

### Continuous predictor

### Interaction effect

## Visualising odds ratios



## Risk ratios (relative risk)

This is **not** the same as an odds ratio.

You will sometimes see people using the two terms interchangeably, but this is wrong.

## Exercises

### Exercise 1

How much more likely is Y outcome in A group vs B group?

For each unit increase in X, how much more likely is Y outcome?


## Summary

::: callout-tip
#### Key points

-   When reporting the results of a logistic regression, it's often helpful to include odds ratios as a measure of effect size
:::

